Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-07-10 patchy-selection-paper.tex
acknowledgements
put in simon's changes and fixed up bibliography

\bibliographystyle{plainnat}

% short title:  Convergent Evolution on Patchy Landscapes
\title 
{
Convergent Evolution During Local Adaptation to Patchy Landscapes
}
% Simon: I think the title only partly reflects the scope of the paper (adaptation by migration being underrepresented). What about "Relative importance of migration and convergent evolution for local adaptation in patchy landscapes" or similar?

@@

Intuitively, convergence is also more likely 
when geographically separated populations adapt to ecologically similar conditions. 
The probability that convergent adaptations arise independently
before adaptations spread between the populations by migration
will be larger if these adaptive alleles are maladapted in intervening environments,
since such adverse conditions can strongly spatially restrict the spread of locally adapted alleles \citep{slatkin1973geneflow}.

@@

Consider a population spread across a landscape, 
within which are patches of habitat to which individuals are (initially) poorly adapted,
surrounded by large areas to which the population is well adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knockout of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a relative fitness
$1-s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
(Here we take ``fitness'' to be the intrinsic growth rate of the type when rare.)
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so are mainly interested in whether the allele
can escape initial loss by drift when rare. 
Therefore, we do not have to specify the fitness of the homozygote; 
only that the dynamics of the allele when rare 
are determined by the fitness of the heterozygote. 
More general dominance will only make small corrections to the dynamics until initial fixation,
with the exception of the recessive case, which we omit.
In other words,
we follow the literature in treating the diploid model as essentially haploid.

@@

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in spatially varying selection models. 
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panmictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}. 
More general work on the mathematically equivalent problem of density-dependent population growth 
has obtained much more general criteria
under which a habitat configuration will support a stable polymorphic equilibrium \citep{cantrell1989diffusive},
and determined the habitat configuration that maximizes the probability of establishment \citep{lou2006minimization}.

@@

Let $p(x)$ be the probability that a new mutant $B$ allele arising at location $x$
relative to the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is symmetric about the edge of the patch.
We examine this more generally via exact calculation of $p(x)$ in Section \ref{apx:establishment_sims}.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, we dissect the process by which an allele transits between the patches,
along the way obtaining other useful information about the genealogy of $B$ alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{haldane1948theory} or \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions (especially regarding scaling) implicit here,
we provide a derivation in Appendix~\ref{apx:eqfreq}; 
as well as justification for the asymptotics in Appendix \ref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see below).

@@

To make the argument, we decompose the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes
whose members are ignored if they enter the original patch.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
This decomposition is illustrated in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(x) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $x$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at displacement $x$ from the original,
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(x) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $x$ ) }
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$.
All terms except the ``outflux of families'' are calculations with spatial branching processes;
and we can solve for this outflux using the established form for $q(r)$ in \eqref{eqn:gestalt_q}.
Note that $\migrate$ reflects the mean number of establishing migrant families per unit time, rather than individuals
-- a naive calculation using $q(x)$ that assumes migrants are independent would overestimate the effective migration rate.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles far from the patch where it is established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $\exp(-s_m)\approx 1-s_m$,
and the variance is $\xi^2$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $e^{-s_m}$ offspring,
the expected family size after time $t$ is $e^{t s_m}$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $e^{-s_m t} = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
then $K_t$ has a limiting distribution: $K_t \dconv K$ 
and $e^{-s_m t}/(1-k_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $e^{-s_m t}$,
and conditional on survival, the family size is given by a fixed distribution.

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary} \citep[see also][]{chauvin1991growing}.
When we condition on survival until $t$, we condition on existence of
at least one ``trunk'' lineage surviving from time $0$ to time $t$ (the long red
lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, nothing distinguishes any of the other reproduction events --
each individual along the trunk lineage must give birth to at least one offspring,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a ``trunk'' that lasts until $t$
which sprouts independent ``branches'' that may or may not survive until $t$.
Concretely, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then this gives the following decomposition:
\begin{align}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is an independent copy of the offspring distribution, conditioned on $W_s \ge 1$.

@@

This suggests that we add in the spatial motion by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}),
whose motion is is approximately Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
We can use these two ingredients, Brownian motion and a branching process, 
to compute the terms of equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate}.

@@

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are
given by a spatial branching process (as depicted in Figure \ref{fig:branching_decomp}).
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Suppose that there is a constant outflux of migrant families from each point $x \in \partial B(r_0)$,
such that the long-term proportion of individuals near a point $x$ in $\partial B(r_0)$ that initiate new migrant families
is $\gamma(x)$.
To make expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} more concrete,
we need to replace ``outflux of families'' by a sum over points at distance $r_0$ from the patch
of the outflux $\gamma$ at that point, multiplied by the relevant second quantity.
By examining the constituent terms,
it will turn out that equation \eqref{eqn:gestalt_migrate} 
is well-approximated by a constant multiple of \eqref{eqn:gestalt_q}.

First consider equation \eqref{eqn:gestalt_q} for the equilibrium frequency.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
and denote by $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
Concretely, it is the density of $\E[\int_0^t Z_t(\cdot) dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $S$, 
gives the expected total number of individuals ever living in $S$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(S) dt \right] = \int_S u(x,y) dy .
\end{align}
% Said another way, $u(x,y)$ is the expected number of offspring of a progenitor at $x$ that will ever be observed near $y$.
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-s_m t} p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-s_m t} p_t(x,y) dt dx .
\end{align}

This form we can now compare to 
the expression \eqref{eqn:gestalt_migrate} for the outflux of successful migrants.
Consider the probability that a migrant family beginning with a single individual at $x$ will ever establish in the new patch,
which we assume is circular, has relatively small area $A$ and is centered at position $y$.
% The probability that this family reaches $A$ at time $t$ is the probability that it survives until $t$, $(1-k_e(t))$,
% multiplied by the probability that the family reaches the patch, given survival.
% If $t$ is large then as in the previous section, $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$.
It would be possible to analyze directly the probability that a given migrant family establishes in the patch,
as in \citet{barton1987establishment};
but we make a simpler approximation, approximating this probability by the chance that the ``trunk'' hits the new patch,
multiplied by the chance that at least one member of the family escapes demographic stochasticity,
successfully establishing in the new patch.
Write $h_A(x,y)$ for the probability that the Brownian trunk
hits the new patch centered at $y$ before the family dies out,
and $f(A)$ for the chance that the family manages to establish in the new patch,
given that it successfully arrives.
As for $q(y)$ above, expression \eqref{eqn:gestalt_migrate} is properly an integral against the outflux $\gamma(x)$:
\begin{align}\label{eqn:migrate_integral}
  \migrate(y) &= \rho \int_{\partial B(r_0)} \gamma(x) h_A(x,y) f(A) dx .
\end{align}

>>>>

% short title:  Convergent Evolution on Patchy Landscapes
\title 
{
Convergent Evolution During Local Adaptation to Patchy Landscapes
}

@@

Intuitively, convergence is also more likely 
when geographically separated populations adapt to ecologically similar conditions. 
The probability that convergent adaptations arise independently
before adaptations spread between the populations by migration
will be larger if these adaptive alleles are maladapted in intervening environments,
since such adverse conditions can strongly restrict the spread of locally adapted alleles \citep{slatkin1973geneflow}.

@@

Consider a population spread across a landscape to which it is generally well adapted,
but within which are patches of habitat to which individuals are (initially) poorly adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knockout of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a relative fitness
$1-s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
(Here we define ``fitness'' to be the intrinsic growth rate of the type when rare.)
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in these intervening regions.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so will focus on whether the allele
can escape initial loss by drift when rare. 
Therefore, we need not specify the fitness of the homozygote; 
only that the dynamics of the allele when rare 
are determined by the fitness of the heterozygote. 
More general dominance will only make small corrections to the dynamics until initial fixation,
with the exception of the recessive case, which we omit.
In other words,
we follow the literature in treating the diploid model as essentially haploid.

@@

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in models of spatially varying selection.
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. 
Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panmictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate $p_e$ by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}. 
More general work on the mathematically equivalent problem of density-dependent population growth 
has obtained much more general criteria
under which a habitat configuration will support a stable polymorphic equilibrium \citep{cantrell1989diffusive},
and determined the habitat configuration that maximizes the probability of establishment \citep{lou2006minimization}.

@@

Let $p(x)$ be the probability that a new mutant $B$ allele that arises at location $x$
relative to the center of the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is symmetric about the edge of the patch.
We examine this more generally via exact calculation of $p(x)$ in Section \ref{apx:establishment_sims}.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, 
study the fine-scale genealogy of alleles that transit between patches,
obtaining along the way other useful information about the genealogy of $B$ alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{haldane1948theory} or \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions (especially regarding scaling) implicit here,
we provide a derivation in Appendix~\ref{apx:eqfreq}, 
as well as justification for the asymptotics in Appendix \ref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see below).

@@

To make the argument, we decompose the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated 
by a collection of independent spatial branching processes
whose members are ignored if they enter the original patch.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
This decomposition is illustrated in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(x) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $x$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate at which \emph{families of} migrant $B$ alleles establish in a new patch at displacement $x$ from the original,
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(x) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $x$ ) }
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$.
All terms except the ``outflux of families'' are calculations with spatial branching processes;
and we can solve for this outflux using the established form for $q(x)$ in \eqref{eqn:gestalt_q}.
% Note that $\migrate$ reflects the mean number of establishing migrant families per unit time, rather than individuals
% -- a naive calculation using $q(x)$ that assumes migrants are independent would overestimate the effective migration rate.

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles 
far from the original patch where it is already established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this as a branching process,
i.e.\ by assuming that each individual
migrates, then reproduces, independently of all other family members.
Migration takes the individual in a random direction
for a random distance whose mean square is $\sigma^2$,
and if the individual arrives to a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring,
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $e^{-s_m}\approx 1-s_m$,
and the variance is $\xi^2$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

Since each individual has on average $e^{-s_m}$ offspring,
the expected family size after time $t$ is $e^{- t s_m}$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $e^{-s_m t} = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed \citep[see][for details]{jagers1975branching},
then $K_t$ has a limiting distribution: $K_t \dconv K$,
and $e^{-s_m t}/(1-k_e(t)) \to \E[K]$ as $t \to \infty$.
In other words, the chance of survival to time $t$ is a constant multiplied by $e^{-s_m t}$,
and conditional on survival, the family size is given by a fixed distribution.

This suggests that families can be well-described by motion of a central ``trunk'',
which dies at a random time whose distribution is given by $k_e(t)$,
and conditional on survival of this trunk, the family is composed of side ``branches'' off of this trunk,
as depicted in Figure~\ref{fig:branching_decomp}.
We can understand this quite concretely, by a method described rigorously by \citet{geiger1999elementary} \citep[see also][]{chauvin1991growing}.
When we condition on survival until $t$, we condition on existence of
at least one trunk lineage surviving from time $0$ to time $t$ (the long red
lineage in Figure \ref{fig:branching_decomp}).
Given this lineage, nothing distinguishes any of the other reproduction events --
each individual along the trunk lineage must give birth to at least one offspring,
and all other reproduction events are unconditioned.
In other words, the genealogy of a family that survives until $t$
can be decomposed into a trunk that lasts until $t$
and that sprouts independent branches that may or may not survive until $t$.
Specifically, if we write $Z_t$ for the number of offspring in the branching process alive at time $t$,
% (so $\P\{Z_t = 0\} = k_e(t)$),
then this gives the following decomposition:
\begin{align}
  \left( Z_t \; \vert \; Z_t>0 \right) \deq 1 + \sum_{s=0}^t \sum_{k=1}^{W_s-1} Z^{(s,k)}_{t-s},
\end{align}
where each $Z^{(s,k)}$ is an independent copy of $Z$ 
that arose from the main branch in generation $s$,
and each $W_s$ is an independent copy of the offspring distribution, conditioned on $W_s \ge 1$.

@@

This suggests that we add in the spatial motion by following only the motion of the ``trunk''
(i.e.\ the red line in figure \ref{fig:branching_decomp}).
Its motion is approximately Brownian, with variance $\sigma^2$ per generation.
(Recall $\sigma$ is the dispersal distance.)
We can use these two ingredients, Brownian motion and a branching process, 
to compute the terms of equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate}.

@@

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are
given by a spatial branching process (as depicted in Figure \ref{fig:branching_decomp}).
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Suppose that there is a constant outflux of migrant families from each point $x \in \partial B(r_0)$,
such that the long-term proportion of individuals near a point $x$ in $\partial B(r_0)$ that initiate new migrant families
is $\gamma(x)$.
To make expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} more concrete,
we need to replace ``outflux of families'' by a sum over points at distance $r_0$ from the patch
of the outflux $\gamma$ at that point.
By examining the constituent terms,
it will turn out that equation \eqref{eqn:gestalt_migrate} 
is well-approximated by a constant multiple of \eqref{eqn:gestalt_q}.

First consider equation \eqref{eqn:gestalt_q} for the equilibrium frequency.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
and denote by $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ at location $y$
can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
It is defined to be the density of $\E[\int_0^t Z_t(\cdot) dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set of locations $S$, 
gives the expected total number of individuals ever living in $S$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(S) dt \right] = \int_S u(x,y) dy .
\end{align}
% Said another way, $u(x,y)$ is the expected number of offspring of a progenitor at $x$ that will ever be observed near $y$.
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-s_m t} p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-s_m t} p_t(x,y) dt dx .
\end{align}

This form we can now compare to 
the expression \eqref{eqn:gestalt_migrate} for the outflux of successful migrants.
Consider the probability that a migrant family beginning with a single individual at $x$ will ever establish in the new patch.
We assume this new patch is circular, 
has relatively small area $A$ and is centered at position $y$.
% The probability that this family reaches $A$ at time $t$ is the probability that it survives until $t$, $(1-k_e(t))$,
% multiplied by the probability that the family reaches the patch, given survival.
% If $t$ is large then as in the previous section, $(1-k_e(t)) \approx e^{-s_m t}/\E[K]$.
It would be possible to analyze this probability directly,
as in \citet{barton1987establishment};
but here we take a simpler route, 
approximating this by the chance that the trunk hits the new patch,
multiplied by the chance that at least one member of the family escapes demographic stochasticity
and successfully establishes in the new patch.
Write $h_A(x,y)$ for the probability that the Brownian trunk
hits the new patch centered at $y$ before the family dies out,
and $f(A)$ for the chance that the family manages to establish in the new patch,
given that it successfully arrives.
As for $q(y)$ above, expression \eqref{eqn:gestalt_migrate} is properly an integral against the outflux $\gamma(x)$:
\begin{align}\label{eqn:migrate_integral}
  \migrate(y) &= \rho \int_{\partial B(r_0)} \gamma(x) h_A(x,y) f(A) dx .
\end{align}
